### 2018-03-07 dissertation/1-intro/chapter1.tex
2018-03-06

As computers grow in power and shrink in size, more aspects of everyday life can be enhanced by adding processing units to common devices.
While many of these applications focus on conveniences, such as home automation \cite{mccole2016} (the collection of connected and smart devices is dubbed the Internet of Things or IoT \cite{morgan2014}), the integration of computers with other objects and devices can also be important to save time and save lives \cite{rti2014}.
One way of achieving this is by adding computers and wireless transmitters to vehicles — such as cars, buses, and trains — so they can share data which may increase traffic efficiency or reduce the chance of accidents \cite{saini2015close}.

In 2013, an estimated 1.25 million people lost their lives due to traffic accidents globally \cite{whotraffic}.
While this number has greatly reduced over the past decades \cite{johnson2010traffic} thanks to better safety features (seat belts, air bags, ABS, etc.) and stronger laws (drunk driving, motorcycle helmets, speed limits, etc.), it may still rise as a major cause of death in the years to come \cite{whofactsheet}, so further actions are necessary.
Furthermore, as the car population increases, congestions consume ever more time of the daily commuter, peaking at over 100 hours per year for the residents of Los Angeles, CA \cite{inrixtraffic}.

Smart vehicles and vehicular networks are ways that technology can aid both of the aforementioned problems.
Through the use of sensors and wireless communications, these vehicles are able to avoid accidents by alerting distracted drivers \cite{lee2004collision}, or by knowing in advance another vehicle's position and speed \cite{hafner2011automated}.
By communicating, they can also collaborate to offer driving and route suggestions, therefore reducing the possibility of traffic jams \cite{knorr2012reducing}.
%These features are possible with the development of a vehicular ad-hoc network (VANET), in which vehicles can quickly share data amongst themselves without the need of a server or an Internet connection.

When dealing with safety or traffic-efficiency applications, it is crucial that network communications occur with low latency (approximately 100 milliseconds \cite{camp2005vehicle}).
Current cellular technology, such as LTE, could be used to connect vehicles to the Internet, but the delay added by the transmission would make safety applications unfeasible or unreliable \cite{mangel2010comparison}.
Cellular connections also have other problems: the connection would require an active subscription with a carrier; the connection depends on available infrastructure; the wireless frequency would be shared with phones and other mobile devices, increasing the possibility of interference and congestion; server-side issues could impact the vehicles' communications.

For these reasons, ad-hoc solutions are preferred over centralized ones.
An ad-hoc network is one that has no reliance on pre-existing infrastructure (such as routers or access points) \cite{wu2004ad}.
Instead, each node is able to communicate directly with others and a routing protocol allows for messages to be forwarded until they reach their destinations.
Every time a node wants to send a message and the recipient is not a direct neighbor, it must choose which nearby node is the most likely one to get the message to its destination.
Routing techniques can use either the network's topology or geographical coordinates \cite{saini2015close} to choose which node should be the next hop.

These issues — the additional safety and efficiency as well as the low-latency communications — can be tackled through the use of a vehicular ad-hoc network (VANET), in which vehicles share data amongst themselves without relying on external devices, an Internet connection or server availability.
Neighboring vehicles can share their position and velocity data at high frequencies, allowing, for example, for autonomous vehicles to plan a platooning approach to traffic \cite{amoozadeh2015platoon}.
In the case of a collision or other event, nearby nodes can broadcast alerts, which other nodes pick up and forward \cite{li2007routing}.
That way, an alert can travel long distances in little time, allowing approaching vehicles to safely slow down or pick alternative routes.

@@

As is the case with most new technologies, VANETs are expected to be a notable target of attacks for a diversity of reasons \cite{isaac2010security}.
A local malicious user might alter the data his or her vehicle broadcasts in order to manipulate traffic conditions, while remote attackers could invade vehicles' computers and obtain partial control of the network  \cite{garip2015congestion}.
These attacks can vary from time-consuming annoyances to life-threatening, so it is important that real-world implementations of vehicular networks are prepared to handle them.

@@

Furthermore, once nodes form their own opinions about others, they can propagate pre-existing trust values when necessary.
For example, if two nodes are not direct neighbors and do not have any pre-existing trust information about each other, they can ask intermediary nodes for their opinions on the other node \cite{wang2009trust}.
The management of trust values (i.e. how one node acquires and updates trust values) and the use of these values to derive further information (such as designating nodes as malicious or not) is called \textit{trust management} \cite{ma2011survey}. 
%The combination of one node's trust of another with the opinions of other nearby nodes form what is called \textit{trust management}.
The effective use of trust management allows for the detection of malicious, misbehaving or faulty nodes in the network, and for such information to be shared amongst the benign participants. 
Throughout this document, trust and trust management may be used interchangeably.

@@

While the detection of incorrect nodes and/or messages is an important aspect of security and safety in vehicular networks, it does not address all of the problems.
Trust solutions are not viable without a solid identity verification scheme, for instance, since nodes would not be able to form trust opinions without being sure of the others' identities (these schemes often use a Public Key Infrastructure \cite{wasef2010complementing}).
They also do not address issues such as driver and passenger privacy when using the facilities of a VANET.
Furthermore, cryptography must be used in order to guarantee that the secrecy of messages is not violated.
Therefore, trust and trust management must be viewed as an important aspect of vehicular ad-hoc networks, but not as a definitive solutions to all of the related concerns.

@@

%While this work emphasizes on the characteristics of VANETs, they are only a subset of the much broader field of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \cite{wang2003complex} \cite{newmannetworks}.
%Trust issues are present in many of these types of networks, although each one requires a specific approach that addresses the peculiarities involved.

%Often, when people hear the term ``network'', they think only of technological examples, such as computer networks.
%Such networks are only a subset of a much larger field of study: that of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \cite{wang2003complex} \cite{newmannetworks}.
%Technological networks, and VANETs by extension, are only a subset of that much larger field of study.
%
%Trust issues and their solutions are an important aspect of the study of complex networks, and of social or technological networks in particular.
%Each different type of network requires a specific approach to trust, according to the features that distinguish it from others.
%It is interesting to review the solutions proposed for social networks and technological networks in general, and then narrow down on how dynamic networks, and VANETs in particular, are handled.
%Despite the differences, there are some concepts that overlap across most types of networks.

This work proposes a method for identifying malicious nodes in a vehicular network, although it may also be viable for other dynamic and decentralized networks.
The solution is based on a previously existing algorithm \cite{vernize2015malicious}, which was limited to centralized networks.
In order to adapt the algorithm to a dynamic and decentralized environment, it is necessary to locate network features that allow for trust relationships to be developed amongst the network members over time, since there is no centralizing agent to store and process the whole network's trust values.
These features are found by tracing analogies to social networks, i.e. identifying how and why two nodes (in the case of VANETs, vehicles) would meet more than once and at a somewhat predictable interval.

>>>>

As computers grow in power and shrink in size, more aspects of everyday life can be enhanced by adding processing units to common devices.
While many of these applications focus on conveniences, such as home automation \citep{mccole2016} (the collection of connected and smart devices is dubbed the Internet of Things or IoT \citep{morgan2014}), the integration of computers with other objects and devices can also be important to save time and save lives \citep{rti2014}.
One way of achieving this is by adding computers and wireless transmitters to vehicles — such as cars, buses, and trains — so they can share data which may increase traffic efficiency or reduce the chance of accidents \citep{saini2015close}.

In 2013, an estimated 1.25 million people lost their lives due to traffic accidents globally \citep{whotraffic}.
While this number has greatly reduced over the past decades \citep{johnson2010traffic} thanks to better safety features (seat belts, air bags, ABS, etc.) and stronger laws (drunk driving, motorcycle helmets, speed limits, etc.), it may still rise as a major cause of death in the years to come \citep{whofactsheet}, so further actions are necessary.
Furthermore, as the car population increases, congestions consume ever more time of the daily commuter, peaking at over 100 hours per year for the residents of Los Angeles, CA \citep{inrixtraffic}.

Smart vehicles and vehicular networks are ways that technology can aid both of the aforementioned problems.
Through the use of sensors and wireless communications, these vehicles are able to avoid accidents by alerting distracted drivers \citep{lee2004collision}, or by knowing in advance another vehicle's position and speed \citep{hafner2011automated}.
By communicating, they can also collaborate to offer driving and route suggestions, therefore reducing the possibility of traffic jams \citep{knorr2012reducing}.
%These features are possible with the development of a vehicular ad-hoc network (VANET), in which vehicles can quickly share data amongst themselves without the need of a server or an Internet connection.

When dealing with safety or traffic-efficiency applications, it is crucial that network communications occur with low latency (approximately 100 milliseconds \citep{camp2005vehicle}).
Current cellular technology, such as LTE, could be used to connect vehicles to the Internet, but the delay added by the transmission would make safety applications unfeasible or unreliable \citep{mangel2010comparison}.
Cellular connections also have other problems: the connection would require an active subscription with a carrier; the connection depends on available infrastructure; the wireless frequency would be shared with phones and other mobile devices, increasing the possibility of interference and congestion; server-side issues could impact the vehicles' communications.

For these reasons, ad-hoc solutions are preferred over centralized ones.
An ad-hoc network is one that has no reliance on pre-existing infrastructure (such as routers or access points) \citep{wu2004ad}.
Instead, each node is able to communicate directly with others and a routing protocol allows for messages to be forwarded until they reach their destinations.
Every time a node wants to send a message and the recipient is not a direct neighbor, it must choose which nearby node is the most likely one to get the message to its destination.
Routing techniques can use either the network's topology or geographical coordinates \citep{saini2015close} to choose which node should be the next hop.

These issues — the additional safety and efficiency as well as the low-latency communications — can be tackled through the use of a vehicular ad-hoc network (VANET), in which vehicles share data amongst themselves without relying on external devices, an Internet connection or server availability.
Neighboring vehicles can share their position and velocity data at high frequencies, allowing, for example, for autonomous vehicles to plan a platooning approach to traffic \citep{amoozadeh2015platoon}.
In the case of a collision or other event, nearby nodes can broadcast alerts, which other nodes pick up and forward \citep{li2007routing}.
That way, an alert can travel long distances in little time, allowing approaching vehicles to safely slow down or pick alternative routes.

@@

As is the case with most new technologies, VANETs are expected to be a notable target of attacks for a diversity of reasons \citep{isaac2010security}.
A local malicious user might alter the data his or her vehicle broadcasts in order to manipulate traffic conditions, while remote attackers could invade vehicles' computers and obtain partial control of the network  \citep{garip2015congestion}.
These attacks can vary from time-consuming annoyances to life-threatening, so it is important that real-world implementations of vehicular networks are prepared to handle them.

@@

Furthermore, once nodes form their own opinions about others, they can propagate pre-existing trust values when necessary.
For example, if two nodes are not direct neighbors and do not have any pre-existing trust information about each other, they can ask intermediary nodes for their opinions on the other node \citep{wang2009trust}.
The management of trust values (i.e. how one node acquires and updates trust values) and the use of these values to derive further information (such as designating nodes as malicious or not) is called \textit{trust management} \citep{ma2011survey}. 
%The combination of one node's trust of another with the opinions of other nearby nodes form what is called \textit{trust management}.
The effective use of trust management allows for the detection of malicious, misbehaving or faulty nodes in the network, and for such information to be shared amongst the benign participants. 
Throughout this document, trust and trust management may be used interchangeably.

@@

While the detection of incorrect nodes and/or messages is an important aspect of security and safety in vehicular networks, it does not address all of the problems.
Trust solutions are not viable without a solid identity verification scheme, for instance, since nodes would not be able to form trust opinions without being sure of the others' identities (these schemes often use a Public Key Infrastructure \citep{wasef2010complementing}).
They also do not address issues such as driver and passenger privacy when using the facilities of a VANET.
Furthermore, cryptography must be used in order to guarantee that the secrecy of messages is not violated.
Therefore, trust and trust management must be viewed as an important aspect of vehicular ad-hoc networks, but not as a definitive solutions to all of the related concerns.

@@

%While this work emphasizes on the characteristics of VANETs, they are only a subset of the much broader field of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \citep{wang2003complex} \citep{newmannetworks}.
%Trust issues are present in many of these types of networks, although each one requires a specific approach that addresses the peculiarities involved.

%Often, when people hear the term ``network'', they think only of technological examples, such as computer networks.
%Such networks are only a subset of a much larger field of study: that of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \citep{wang2003complex} \citep{newmannetworks}.
%Technological networks, and VANETs by extension, are only a subset of that much larger field of study.
%
%Trust issues and their solutions are an important aspect of the study of complex networks, and of social or technological networks in particular.
%Each different type of network requires a specific approach to trust, according to the features that distinguish it from others.
%It is interesting to review the solutions proposed for social networks and technological networks in general, and then narrow down on how dynamic networks, and VANETs in particular, are handled.
%Despite the differences, there are some concepts that overlap across most types of networks.

This work proposes a method for identifying malicious nodes in a vehicular network, although it may also be viable for other dynamic and decentralized networks.
The solution is based on a previously existing algorithm \citep{vernize2015malicious}, which was limited to centralized networks.
In order to adapt the algorithm to a dynamic and decentralized environment, it is necessary to locate network features that allow for trust relationships to be developed amongst the network members over time, since there is no centralizing agent to store and process the whole network's trust values.
These features are found by tracing analogies to social networks, i.e. identifying how and why two nodes (in the case of VANETs, vehicles) would meet more than once and at a somewhat predictable interval.

### 2018-03-07 dissertation/2-complexnetworks/chapter2.tex
2018-03-06

%Often, when people hear the term ``network'', they think only of technological examples, such as computer networks.
%Such networks are only a subset of a much larger field of study: that of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \cite{wang2003complex} \cite{newmannetworks}.
%Technological networks, and VANETs by extension, are only a subset of that much larger field of study.
%
%Trust issues and their solutions are an important aspect of the study of complex networks, and of social or technological networks in particular.
%Each different type of network requires a specific approach to trust, according to the features that distinguish it from others.
%It is interesting to review the solutions proposed for social networks and technological networks in general, and then narrow down on how dynamic networks, and VANETs in particular, are handled.
%Despite the differences, there are some concepts that overlap across most types of networks.

Complex networks can describe many systems which are observed in nature and society through a collection of \textit{vertices} (or \textit{nodes}) and \textit{edges} \cite{newmannetworks}.
They can be comprised of palpable components (such as computers and cables), somewhat abstract entities (such as the World Wide Web's collection of webpages and URLs), or both (like the people and relationships that form a social network).
The study of networks is tied to graph theory, since graphs are a useful way to generate models of networks, and therefore many concepts of the two fields overlap.
Mathematical, computational and statistical concepts developed for graphs can be translated to be useful for a variety of different types of networks, like, for example, finding articulation vertices in a graph to locate bandwidth bottlenecks in a computer network.

Complex networks are generally divided into four categories \cite{newmannetworks}:

@@

Trust is a concept studied in fields such as psychology and economics, with specific definitions.
In complex networks, under the perspective of computer science, it is a measure of one entity's confidence that another will behave properly and provide valid and/or meaningful data \cite{sherchan2013survey}.
What follows is the basis of how a network can be modelled using a graph and how a trust model can be applied to it.

@@

\section{Trust in Social Networks}
\label{section:trustsocial}
There are two types of social networks: real-world ones formed by relationships between people, and online ones that attempt to abstract the former into a digital environment.
Examples of the first one are all around, present in any family, workplace, school or group of friends \cite{newmannetworks}.
Online social networks started by connecting people who already knew each other and giving them an additional form of interaction (analogous to what telephones and email did before), but, today, it is not unusual for people to form relationships with others whom they have only met online.

In a traditional social network, it is simple to perceive how trust is relevant and how it works, since trust relationships between people are used on a daily basis to make decisions.
When adapted to a digital environment, these social relationships can be used to automatically increase the relevance of certain information.
For instance, upon reading an online review for a certain product, a user will be more likely to accept the review's conclusion if it was written by a close friend than if it were written by a stranger.
Social trust is a way of estimating how much a certain recommendation will lead to a positive outcome \cite{golbeck2006inferring}.

@@

Social networks also have the property of carrying trust from one relationship to another:
information shared by a close friend of a person might be considered almost as trustworthy as some collected by the person him or herself.
Therefore, it is possible to model social trust relationships as a graph, in which nodes represent people and edges represent a certain degree of trust \cite{newmannetworks}.
Expanding on that property, there is the concept friends of friends \cite{boissevain1974friends}.
If, for example, nodes $A$ and $B$ have mutual trust and are considered friends, then it is reasonable to assume that some of $A$'s trust for $B$ carries over to other nodes that enjoy mutual trust with $B$.
In other words, a friend of a friend can be considered more trustworthy than the average stranger.
This property is similar, although not identical, to transitivity, since trust is diminished for each extra step an origin node needs to reach a destination, and there is also the possibility that one node distrusts another even if they share a mutual friend.
Naturally, social trust is not commutative ($A$ trusting $B$ does not imply that $B$ trusts $A$).

@@

Another important aspect of trust in ad-hoc networks is that information is, generally, uncertain and incomplete \cite{baras2005cooperation}.
That is, since nodes form their own model and opinions of the surrounding network, it is unlikely that this data will be certain and accurate with reality.
For this reason, it is also possible to use data gathered by neighboring nodes to complement the model.
This data itself is subject to the trust evaluation of the neighbors, but it is crucial to better approximate the trust values of other nodes.
Incompleteness is an inherit trace of MANETs, since it is entirely possible for nodes to be too distant to communicate, and only occasionally come into contact.

@@

There are few examples of MANETs implemented for consumer devices.
Two examples are networks created to quickly share data between devices using Wi-Fi or Bluetooth \cite{krochmal2014discovery}, or ones that allow for multiplayer gaming sessions amongst multiple nearby devices \cite{sasaki2011wireless}.
In both cases, there is no need for a complicated system-level trust model, since those activities involve active participation from the users wielding the device (that is, the user chooses whether or not to communicate with other devices); rather, the trust relationship occurs socially amongst the users themselves.

>>>>

%Often, when people hear the term ``network'', they think only of technological examples, such as computer networks.
%Such networks are only a subset of a much larger field of study: that of complex networks.
%The study of complex networks dates back decades, including analysis of biological or social networks \citep{wang2003complex} \citep{newmannetworks}.
%Technological networks, and VANETs by extension, are only a subset of that much larger field of study.
%
%Trust issues and their solutions are an important aspect of the study of complex networks, and of social or technological networks in particular.
%Each different type of network requires a specific approach to trust, according to the features that distinguish it from others.
%It is interesting to review the solutions proposed for social networks and technological networks in general, and then narrow down on how dynamic networks, and VANETs in particular, are handled.
%Despite the differences, there are some concepts that overlap across most types of networks.

Complex networks can describe many systems which are observed in nature and society through a collection of \textit{vertices} (or \textit{nodes}) and \textit{edges} \citep{newmannetworks}.
They can be comprised of palpable components (such as computers and cables), somewhat abstract entities (such as the World Wide Web's collection of webpages and URLs), or both (like the people and relationships that form a social network).
The study of networks is tied to graph theory, since graphs are a useful way to generate models of networks, and therefore many concepts of the two fields overlap.
Mathematical, computational and statistical concepts developed for graphs can be translated to be useful for a variety of different types of networks, like, for example, finding articulation vertices in a graph to locate bandwidth bottlenecks in a computer network.

Complex networks are generally divided into four categories \citep{newmannetworks}:

@@

Trust is a concept studied in fields such as psychology and economics, with specific definitions.
In complex networks, under the perspective of computer science, it is a measure of one entity's confidence that another will behave properly and provide valid and/or meaningful data \citep{sherchan2013survey}.
What follows is the basis of how a network can be modelled using a graph and how a trust model can be applied to it.

@@

\section{Trust in Social Networks}
\label{section:trustsocial}
There are two types of social networks: real-world ones formed by relationships between people, and online ones that attempt to abstract the former into a digital environment.
Examples of the first one are all around, present in any family, workplace, school or group of friends \citep{newmannetworks}.
Online social networks started by connecting people who already knew each other and giving them an additional form of interaction (analogous to what telephones and email did before), but, today, it is not unusual for people to form relationships with others whom they have only met online.

In a traditional social network, it is simple to perceive how trust is relevant and how it works, since trust relationships between people are used on a daily basis to make decisions.
When adapted to a digital environment, these social relationships can be used to automatically increase the relevance of certain information.
For instance, upon reading an online review for a certain product, a user will be more likely to accept the review's conclusion if it was written by a close friend than if it were written by a stranger.
Social trust is a way of estimating how much a certain recommendation will lead to a positive outcome \citep{golbeck2006inferring}.

@@

Social networks also have the property of carrying trust from one relationship to another:
information shared by a close friend of a person might be considered almost as trustworthy as some collected by the person him or herself.
Therefore, it is possible to model social trust relationships as a graph, in which nodes represent people and edges represent a certain degree of trust \citep{newmannetworks}.
Expanding on that property, there is the concept friends of friends \citep{boissevain1974friends}.
If, for example, nodes $A$ and $B$ have mutual trust and are considered friends, then it is reasonable to assume that some of $A$'s trust for $B$ carries over to other nodes that enjoy mutual trust with $B$.
In other words, a friend of a friend can be considered more trustworthy than the average stranger.
This property is similar, although not identical, to transitivity, since trust is diminished for each extra step an origin node needs to reach a destination, and there is also the possibility that one node distrusts another even if they share a mutual friend.
Naturally, social trust is not commutative ($A$ trusting $B$ does not imply that $B$ trusts $A$).

@@

Another important aspect of trust in ad-hoc networks is that information is, generally, uncertain and incomplete \citep{baras2005cooperation}.
That is, since nodes form their own model and opinions of the surrounding network, it is unlikely that this data will be certain and accurate with reality.
For this reason, it is also possible to use data gathered by neighboring nodes to complement the model.
This data itself is subject to the trust evaluation of the neighbors, but it is crucial to better approximate the trust values of other nodes.
Incompleteness is an inherit trace of MANETs, since it is entirely possible for nodes to be too distant to communicate, and only occasionally come into contact.

@@

There are few examples of MANETs implemented for consumer devices.
Two examples are networks created to quickly share data between devices using Wi-Fi or Bluetooth \citep{krochmal2014discovery}, or ones that allow for multiplayer gaming sessions amongst multiple nearby devices \citep{sasaki2011wireless}.
In both cases, there is no need for a complicated system-level trust model, since those activities involve active participation from the users wielding the device (that is, the user chooses whether or not to communicate with other devices); rather, the trust relationship occurs socially amongst the users themselves.
